Adaptive evolution with aneuploidy and mutation

### 2021-06-17 ms/ms.tex
add MP, AHY as authors
started revision

@@

% Title page
% Chromosomal duplication is a transient evolutionary solution to stress
\title{Clonal interference between aneuploidy and mutation}

@@

\author[1]{Ilia Kohanovski}
%\author[b,c]{Avihu H. Yona}
%\author[d]{Orna Dahan}
%\author[d]{Yitzhak Pilpel}
\author[1,2,*]{Yoav Ram}

\affil[1]{School of Computer Sciences, IDC Herzliya, Herzliya, Israel}
\affil[2]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv, Israel}
%\affil[b]{Department of Physics, Massachusetts Institute of Technology, Cambridge, MA 02139, USA}{
%\affil[b]{Department of Biological Engineering, Massachusetts Institute of Technology, Cambridge, MA 02139, USA}
%\affil[d]{Department of Molecular Genetics, Weizmann Institute of Science, Rehovot 76100, Israel}
\affil[*]{Corresponding author: yoav@yoavram.com}

@@

% Abstract
\begin{abstract}
Aneuploidy is common in eukaryotes, often leading to decreased cell growth and fitness. However, evidence from yeast and fungi, as well as tumour cells, suggests that aneuploidy can be beneficial under stressful conditions and  lead to elevated growth rates and to adaptation. Thus, it is crucial to develop a quantitative theory for the role of aneuploidy in adaptive evolution. Here, we analyse results from experiments in which \yeast adapted to heat stress. The experimental population first acquired chromosome duplications, only to later revert back to a euploid state. We use several evolutionary models within a approximate Bayesian computation framework to show that clonal interference between chromosome duplications and beneficial mutations explains the experimental results. Our results suggest that aneuploidy can both accelerate and deccelerate adaptation in a non-intuitive manner, creating an evolutionary conflict between the individual and the population and further complicating the process of adaptation in cell populations.
\end{abstract}

@@

\paragraph*{Aneuploidy is common in eukaryotes.}
Aneuploidy is an imbalance in the number of chromosomes in the cell: an incorrect karyotype.
Recent evidence suggests aneuploidy is very common in eukaryotes, e.g. animals~\citep{Santaguida2015review, Naylor2016, Bakhoum2017}, and fungi~\citep{Pavelka2010, Zhu2016, Robbins2017, Todd2017}.
Aneuploidy has been implicated in cancer formation and progression~\citep{Boveri2008, Schvartzman2010}: 
90\% of solid tumours and 50\% of blood cancers are aneuploid~\citep{Santaguida2015review}.
Aneuploidy is also linked to the emergence of drug resistance~\citep{Selmecki2009} and virulence~\citep{Moller2018} in fungal pathogens, which are under-studied~\citep{Rodrigues2018} despite infecting close to a billion people per year, causing serious infections and significant morbidity in >150 million people per year and killing >1.5 million people per year~\citep{Selmecki2009, Rodrigues2018}.
In addition, aneuploidy is common in protozoan pathogens of the Leishmania genus, a major global health concern~\citep{Mannaert2012}.

\paragraph*{Aneuploidy is generally deleterious.}
The molecular and genetic mechanisms involved in aneuploidy have been explored in the past decade~\citep{Musacchio2007, Sheltzer2011, Chen2012b, Rancati2013, Gerstein2015, Shor2015}.
Experiments with human and mouse embryos found that aneuploidy is usually lethal;
it is also associated with developmental defects and lethality in other multicellular organisms~\citep{Sheltzer2011}. For example, aneuploid mouse embryonic cells grow slower than euploid cells~\citep{Williams2008}.
Similarly, in unicellular eukaryotes growing in benign conditions, aneuploidy usually leads to slower growth and decreased overall fitness~\citep{Niwa2006, Torres2007, Pavelka2010, Sheltzer2011, Kasuga2016}, in part due to proteotoxic stress caused by increased expression in aneuploid cells~\citep{Pavelka2010, Santaguida2015, Zhu2018} and hypo-osmotic-like stress~\citep{Tsai2019}.

@@

\paragraph*{Transient adaptive solution.} 
Aneuploidy differs from mutation due to its distinct properties. 
Chromosome duplication usually occurs more often than mutation and on average produces larger fitness effects.
Yet, because it affects many genes on a whole chromosome or a chromosome fragment, aneuploidy also carries fitness costs.
Thus, aneuploidy can be a \emph{transient adaptive solution}: it can rapidly occur and fix in the population under stressful conditions, and can be rapidly lost when the cost outweighs the benefit -- when stress is removed or after beneficial mutations occur.
Experimental evidence of such a transient role of aneuploidy was demonstrated by \citet{Yona2012}, who evolved populations of \yeast under strong heat or pH stress.
In these experiments, the populations quickly adapted to the stress, and this adaptation was determined to be due to chromosome duplications.
Later on, the populations reverted back to an euploid state, while remaining adapted to the stress and accumulating multiple mutations.
However, under gradual heat stress, aneuploidy was not observed.
\citet{Yona2012} concluded that aneuploidy serves as a transient adaptive solution, or a ``quick fix'', which is expected to facilitate adaptation.

\paragraph*{The present study.}
Here, we develop a series of computational evolutionary models that include the effects of natural selection, genetic drift, aneuploidy and mutation to examine the role of aneuploidy in adaptation.
These models follow a population of cells characterised by both their ploidy and their genotype.
We fit these models to the empirical results of \citet{Yona2012} using approximate Bayesian computation.
These model fits allow us to infer model parameters, including selection coefficients and transition rates, and to perform model selection between different models, thereby testing different hypotheses about the evolutionary process.
We find that % TODO CHECK!
the aneuploidy rate is several orders of magnitude higher than the mutation rate; that a simple model of clonal interference is not enough to explain the transience of aneuploidy; and that aneuploidy rescues the population, albeit at the price of delaying further adaptation.

@@

\paragraph*{Evolutionary Models.} To examine the role of aneuploidy in adaptation, two models were developed - \emph{single-mutant model} and \emph{multiple-locus model}. 
Both are based on the Wright-Fisher model~\citep{Tenaillon1999} and include the effects of natural selection, genetic drift, aneuploidy, and mutation. Only beneficial mutations assumed, 
thus neglecting the effects of deleterious and neutral mutations. The \emph{single-mutant model} allows for a single mutation (on specific locus or target loci), while the 
\emph{multiple-locus model} allows for multiple mutations. Both models allow for a single aneuploid karyotype (e.g. chromosome III duplication). (\autoref{fig:models})

\begin{figure}[b!]
  \centering
  \begin{subfigure}[t]{0.5\textwidth}
      \centering
      \includegraphics[height=1.8in]{../figures/Fig1-A.pdf}
      \caption{
        \textbf{A. Single-mutant}
        }
      \label{fig:model1}
  \end{subfigure}%
  ~ 
  \begin{subfigure}[t]{0.5\textwidth}
      \centering
      \includegraphics[height=1.8in]{../figures/Fig1-B.pdf}
      \caption{
        \textbf{B. Multiple-locus}
      }
      \label{fig:model2}
  \end{subfigure}
  \caption{
    \textbf{Model Schemes.} (A) The population with four possible genotypes and four possible transitions from one genotype to another with the transition rates $\mu$, $\alpha$, 
    $\alpha_l$, $k\mu$. \emph{2n}: euploid wildtype – the initial genotype; \emph{2n*}: euploid mutant, standard karyotype with one specific beneficial mutation; 
    \emph{2n+1}: aneuploid wildtype, with an extra chromosome, e.g. following chromosome duplication; \emph{2n+1*}: 
    aneuploid mutant, with both the beneficial mutation and the extra chromosome. Overall there are two possible paths to reach \emph{2n*} from \emph{2n}.
    (B) Each genotype can have zero or more beneficial mutations and zero or one extra chromosomes. $\mu$, mutation rate; $\alpha$, aneuploidy gain rate; $\alpha_l$, aneuploidy loss rate
  }
  \label{fig:models}
\end{figure}

\paragraph*{Single-mutant model.} It contains four predefined genotypes whose frequencies change during generations. \emph{2n}: euploid wildtype - the initial genotype; 
\emph{2n*}: euploid mutant, standard karyotype with one specific beneficial mutation; \emph{2n+1}: aneuploid wildtype, with an extra chromosome, e.g. following chromosome duplication; 
\emph{2n+1*}: aneuploid mutant, with both the beneficial mutation and the extra chromosome.

The parameters of the model are: 
$N$, the population size, which is constant in all generations; $\mu$, the mutation rate - the transition rate from \emph{2n} to \emph{2n*}, i.e. the frequency at which cells with genotype \emph{2n*}
are created from cells with genotype \emph{2n}; $\alpha$, the aneuploidy gain rate - the transition rate from \emph{2n} to \emph{2n+1}; 
$\alpha_l$, the aneuploidy loss rate - the transition rate from \emph{2n+1*} to \emph{2n*}; $k$, the coefficient, such that $k\cdot\mu$ is the transition rate from \emph{2n+1} to \emph{2n+1*}, 
where on default $k$ is equals to $33/32$ (euploid \emph{2n} genotyped \emph{S. cerevisiae} contains 32 chromosomes, therefore, assuming each chromosome have equal probability to have the specific 
beneficial mutation in, aneuploid \emph{2n+1} genotype should have 33/32 times the mutation rate); $s$, the mutation selection coefficient; $s_b$, fitness benefit of aneuploidy; 
$c$, fitness cost of aneuploidy, where $0<c<1$. The relative finesses of genotypes are:
\begin{equation} \label{eq:single-w} \begin{aligned}
w_{2n} & = 1 \\
w_{2n+1} & = 1-c+s_b \\
w_{2n+1*} & = (1+s)(1-c)+s_b \\
w_{2n*} & = (1+s) \;
\end{aligned}
\end{equation}

\pagebreak

The model follows Wright-Fisher \citep{Tenaillon1999} steps. The first generation is initialized with $N$ cells of the genotype \emph{2n}. At each generation, the frequency $f_g$ of 
each genotype $g$ is updated:
\begin{enumerate}
  \item Selection step: 
    \begin{equation} \label{eq:single-selection} 
      f_g\gets{\dfrac{f_g w_g}{\sum_{j}f_j w_j}}
      %  j\in\{2n, 2n+1, 2n+1*, 2n*\} 
    \end{equation}
  \item Mutation gain and aneuploidy gain/loss step:
    \begin{equation} \label{eq:single-mutation} \begin{aligned}
      f_{2n} & \gets{f_{2n} - \alpha f_{2n} - \mu f_{2n}}  \\
      f_{2n+1} & \gets{f_{2n+1} - k\mu f_{2n+1} + \alpha f_{2n}}  \\
      f_{2n+1*} & \gets{f_{2n+1*} - \alpha_l f_{2n+1*} + k\mu f_{2n+1}}  \\
      f_{2n*} & \gets{\mu f_{2n} + \alpha_l f_{2n+1}}  \;
    \end{aligned}
    \end{equation}
  \item Random drift step: draw the frequencies of all genotypes from a multinomial distribution
    \begin{equation} \label{eq:single-mutation} 
      f \sim Mult(N, (f_{2n},f_{2n+1},f_{2n+1*},f_{2n*}))
    \end{equation}
\end{enumerate}

\paragraph*{Multiple-locus model.} It expands the \emph{single-mutant model} by allowing for any number of beneficial mutations. For each mutation $i$, selection coefficient $s_i$ is independently drawn from an exponential distribution with expected value $s$. The parameters of the model are $\mu$, the mutation rate; $\alpha$, the aneuploidy gain rate; $\alpha_l$, the aneuploidy loss rate; $s$, the parameter of the exponential distribution; $s_b$, fitness benefit of aneuploidy; $c$, fitness cost of aneuploidy, where $0<c<1$. In this model, the population may include any number of genotypes that contain zero or more mutations and zero or one extra chromosomes. The fitness $w_g$ of genotype $g$ is:

\begin{equation} \label{eq:multiple-fitness}
w_g= 
\begin{cases}
   1, & \text{wildtype} \\
	1-c+s_b, & \text{aneuploid wildtype} \\
	\prod_{\text{i is mutation in g}}{(1+s_i)}, & \text{euploid mutant} \\ 
	s_b+(1-c)\prod_{\text{i is mutation in g}}{(1+s_i)}, & \text{aneuploid mutant}
\end{cases}
\end{equation}

@@

\paragraph*{Aneuploidy and mutation: \textit{same-same but different}.}
The published data indicate that, like mutation, aneuploidy can be both deleterious and beneficial~\citep{Pavelka2010, Sheltzer2011}.
Nevertheless, there are important and fundamental differences between adaptation by aneuploidy
and adaptation by beneficial mutations~\citep{Yona2015}, which make aneuploidy a unique mechanism for generating genetic
variation.
First, the aneuploidy rate (i.e. the frequency of mis-segregation events) is significantly higher than the
mutation rate~\citep{Santaguida2015review}.
Thus, everything else being equal, adaptation by aneuploidy will be faster and more frequent.
Second, fitness effects of aneuploidy are larger than those of the majority of mutations, on average, and are rarely
neutral~\citep{Pavelka2010, Yona2012, Sunshine2015}, allowing selection to quickly sort deleterious and beneficial genotypes.
Third, the number of different karyotypes is considerably smaller than the number of different genotypes, and different karyotypes are likely to have different phenotypes~\citep{Pavelka2010}.
Therefore, exploration of the phenotype space by aneuploidy requires smaller populations and a shorter time span.
Fourth, aneuploidy is a reversible state, as the rate of chromosome loss is high and the cost of aneuploidy is significant~\citep{Niwa2006}.
Indeed, aneuploidy often provides a transient solution: under short-term stress conditions, aneuploidy reverts (chromosome number returns to normal) when the stress subsides; under long-term stress conditions, aneuploidy reverts when refined solutions, generated by beneficial mutations, take over~\citep{Yona2012}.
Finally, aneuploidy results in increased genome instability, potentially increasing genetic variation by a positive feedback loop~\citep{Rancati2013, Bouchonville2009, Zhu2012}, while also increasing its own transience.

@@

\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Lilach Hadany, Judith Berman, David Gresham, and Martin Kupiec for discussions and comments.
This work was supported in part by 
the Israel Science Foundation 552/19 and Minerva
Stiftung Center for Lab Evolution (YR).
}

>>>>

% genotype commands
\newcommand{\euwt}{\emph{2n}}
\newcommand{\anwt}{\emph{2n+1}}
\newcommand{\eumt}{\emph{2n*}}
\newcommand{\anmt}{\emph{2n+1*}}

@@

% Title page
% Chromosomal duplication is a transient evolutionary solution to stress
%\title{Clonal interference between aneuploidy and mutation}
\title{Adaptive evolution with aneuploidy and mutation}

@@

\author[1,*]{Ilia Kohanovski}
\author[2,*]{Martin Pontz}
\author[3]{Avihu H. Yona}
%\author[d]{Orna Dahan}
%\author[d]{Yitzhak Pilpel}
\author[1,2,$\dagger$]{Yoav Ram}

\affil[1]{School of Computer Sciences, IDC Herzliya, Herzliya, Israel}
\affil[2]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv, Israel}
\affil[3]{Institute of Biochemistry, Food Science and Nutrition,
Robert H. Smith Faculty of Agriculture, Food and Environment,
The Hebrew University of Jerusalem, Israel}
%\affil[d]{Department of Molecular Genetics, Weizmann Institute of Science, Rehovot 76100, Israel}
\affil[*]{These authors contributed equally to this work}
\affil[$\dagger$]{Corresponding author: yoav@yoavram.com}

@@

% Abstract
\begin{abstract}
Aneuploidy is common in eukaryotes, often leading to decreased cell growth and fitness. However, evidence from yeast and fungi, as well as human tumour cells, suggests that aneuploidy can be beneficial under stressful conditions and lead to elevated growth rates and adaptation.
Because aneuploidy differs from mutation in rate, expected effect, and reversibility, it is crucial to develop a quantitative theory for the role of aneuploidy in adaptive evolution.
Here, we develop evolutionary models for adaptive evolution with both mutation and aneuploidy. 
These models are used within an approximate Bayesian computation framework to estimate the formation rate and fitness effect of aneuploidy and mutation from empirical results of experiments in which \yeast adapted to heat stress. The experimental population first acquired chromosome duplications, only to later revert back to a euploid state.
We also analyze our models to estimate the effect of the aneuploidy and mutation rates on the expected adaptation time and the probability for adaptation via aneuploidy.
Our results suggest that aneuploidy is a transient adaptive solution, which can decelerate adaptation in a non-intuitive manner. By creating an evolutionary conflict between the individual and the population, aneuploidy further complicates  the process of adaptation in cell populations.
\end{abstract}

@@

\paragraph*{Aneuploidy is common in eukaryotes.}
Aneuploidy is an imbalance in the number of chromosomes in the cell: an incorrect karyotype.
Evidence suggests aneuploidy is very common in eukaryotes, e.g. animals~\citep{Santaguida2015review, Naylor2016, Bakhoum2017}, and fungi~\citep{Pavelka2010, Zhu2016, Robbins2017, Todd2017}.
Aneuploidy has been implicated in cancer formation and progression~\citep{Boveri2008, Schvartzman2010}: 
90\% of solid tumours and 50\% of blood cancers are aneuploid~\citep{Santaguida2015review}.
Aneuploidy is also linked to the emergence of drug resistance~\citep{Selmecki2009} and virulence~\citep{Moller2018} in fungal pathogens, which are under-studied~\citep{Rodrigues2018} despite infecting close to a billion people per year, causing serious infections and significant morbidity in >150 million people per year and killing >1.5 million people per year~\citep{Selmecki2009, Rodrigues2018}.
In addition, aneuploidy is common in protozoan pathogens of the Leishmania genus, a major global health concern~\citep{Mannaert2012}.

\paragraph*{Aneuploidy is generally deleterious.}
The molecular and genetic mechanisms involved in aneuploidy have been explored~\citep{Musacchio2007, Sheltzer2011, Chen2012b, Rancati2013, Gerstein2015, Shor2015}.
Experiments with human and mouse embryos found that aneuploidy is usually lethal.
It is also associated with developmental defects and lethality in other multicellular organisms~\citep{Sheltzer2011}. For example, aneuploid mouse embryonic cells grow slower than euploid cells~\citep{Williams2008}.
Similarly, in unicellular eukaryotes growing in benign conditions, aneuploidy usually leads to slower growth and decreased overall fitness~\citep{Niwa2006, Torres2007, Pavelka2010, Sheltzer2011, Kasuga2016}, in part due to proteotoxic stress caused by increased expression in aneuploid cells~\citep{Pavelka2010, Santaguida2015, Zhu2018} and hypo-osmotic-like stress~\citep{Tsai2019}.

@@

\paragraph*{Transient adaptive solution.} 
Aneuploidy differs from mutation due to its distinct properties. 
Chromosome duplication usually occurs more often than mutation and on average produces larger fitness effects.
Yet, because it affects many genes on a whole chromosome or a chromosome fragment, aneuploidy also carries fitness costs.
Thus, aneuploidy can be a \emph{transient adaptive solution}: it can rapidly occur and fix in the population under stressful conditions, and can be rapidly lost when the cost outweighs the benefit---when stress is removed or after beneficial mutations occur.
Experimental evidence of such a transient role of aneuploidy was demonstrated by \citet{Yona2012}. They evolved populations of \yeast under strong heat or pH stress.
In these experiments, the populations adapted to the stress within 450 generations, and this adaptation was determined to be due to chromosome duplications.
Much later, after more than 1500 generations, the populations reverted back to an euploid state, while remaining adapted to the stress and accumulating multiple mutations.
However, under gradual heat stress, aneuploidy was not observed.
\citet{Yona2012} concluded that aneuploidy serves as a transient adaptive solution, or a ``quick fix'', which is expected to facilitate adaptation.

\paragraph*{The present study.}
Here, we develop evolutionary-genetic models that include the effects of natural selection, genetic drift, aneuploidy, and mutation to examine the role of aneuploidy in adaptive evolution.
These models follow a population of cells characterised by both their ploidy and their genotype.
We fit these models to the experimental results of \citet{Yona2012} using an \emph{approximate Bayesian computation} framework~\citep{Sisson2009, Klinger2018} to infer model parameters, including selection coefficients and rates of aneuploidy and mutation, and to perform model selection between different models, thereby testing different hypotheses about the evolutionary process.
Furthermore, we analyze these models to estimate the effects of parameters on the adaptation time and the probability for adaptation via aneuploidy.
We find that % TODO CHECK!
the aneuploidy rate is several orders of magnitude higher than the mutation rate; that a simple model of clonal interference is not enough to explain the transience of aneuploidy; % what does that mean?
and that aneuploidy is likely to fix in the population and increase its mean fitness, but at the price of delaying further adaptation.

@@

\begin{figure}[b!]
  \centering
  \begin{subfigure}[t]{0.5\textwidth}
      \caption{
        \textbf{A. Single-locus model}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/Fig1-A.pdf}      
      \label{fig:model1}
  \end{subfigure}%
  \\
  \begin{subfigure}[t]{0.5\textwidth}
  	  \caption{
        \textbf{B. Multi-locus model}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/Fig1-B.pdf}      
      \label{fig:model2}
  \end{subfigure}
  \caption{
    \textbf{Model illustrations.}
    \textbf{(A)} In the single-locus model, the four genotypes are: euploid wild-type, \euwt; euploid mutant, \eumt; aneuploid wild-type, \anwt; and aneuploid mutant, \anmt.
    Overall there are two possible trajectories from \euwt to \eumt.
    \textbf{(B)} In the multi-locus model, each genotype is characterized by its karyotype, \euwt or \anwt, and the number of accumulated beneficial mutations, denoted by stars. 
    In both panels arrows denote transitions between genotypes, with transitions rates: $\mu$, mutation rate; $\alpha$, aneuploidy gain rate; $\alpha_l$, aneuploidy loss rate.
  }
  \label{fig:models}
\end{figure}

\paragraph*{Evolutionary Models.}
We developed two models: a single-locus model and a multi-locus model. 
Both models are based on the Wright-Fisher model~\citep{Otto2007}, assuming a constant population size $N$ with non-overlapping generations and including the effects of natural selection, genetic drift, aneuploidy, and mutation.
We focus on beneficial mutations, neglecting the effects of deleterious and neutral mutations. Both models allow for a single aneuploid karyotype (e.g., chromosome III duplication).
While the single-locus model allows for only a single mutation to occur, the
multi-locus model allows for multiple mutations to accumulate in the genome~(\autoref{fig:models}).

In the following we describe the

\paragraph*{Single-locus model.}
This model follows four genotypes: euploid wild-type, \euwt, the initial genotype; 
euploid mutant, \eumt, with the standard karyotype and a single beneficial mutation; 
aneuploid wild-type, \anwt, with an extra chromosome, e.g. following chromosome duplication; and
aneuploid mutant, \anmt, with and extra chromosome and a beneficial mutation.

Beneficial mutations from \euwt\; to \eumt\; occur with probability $\mu$, the mutation rate, and from \anwt\; to \anmt\; with probability $\tau \mu$, where $\tau$ is the fold-change in the rate of beneficial mutations in aneuploid cells. By default, we assume $\tau=33/32$, as the wild-type \yeast strains in the experiments by \citep{Yona2012} are diploid, with 32 chromosomes, and the aneuploid strains are trisomic, with 33 chromosomes.
Aneuploidy is formed by chromosome missegregation, so that cells transition from \euwt\; to \anwt\; or from \eumt\; to \anmt\; with probability $\delta$, the aneuploidy gain rate.
Aneuploidy is lost, transitioning cells from \anwt\; to \euwt\; or from \anmt\; to \eumt\; with probability $\delta_L$, the aneuploidy loss rate.
The fitness values of the four genotypes are given by
\begin{equation} \label{eq:fitness-single}
\begin{aligned}
&w_{2n}  = 1 \;\;&
&w_{2n+1}  = 1-c+\sigma \;\;&
\\
&w_{2n+1^*}  = (1-c)(1+s)+\sigma \;\;&
&w_{2n^*}  = 1+s \;,\;&
\end{aligned}
\end{equation}
such that $s \ge 0$ is the selection coefficient of a beneficial mutation;
$\sigma \ge 0$ is the selection coefficient of aneuploidy; 
and $0 \le c \le 1$ is the fitness cost of aneuploidy.

The first generation is initialized with $N$ cells with genotype \euwt.
The effect of natural selection on the frequency $f_i$ of genotype $i = 2n, 2n+1, 2n+1^*, \text{or } 2n^*$ is given by
    \begin{equation} \label{eq:selection-single} 
      f^s_i = \frac{f_i w_i}{\bar{w}} \;,
    \end{equation}
where $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
The effect of mutation and aneuploidy on genotype frequencies is given by
    \begin{equation} \label{eq:mutation-aneuploidy-single}
    \begin{aligned}
      f^m_{2n} & = {f^s_{2n} - \delta f^s_{2n} - \mu f^s_{2n}}  \;,\\
      f^m_{2n+1} & = {f^s_{2n+1} - \tau \mu f^s_{2n+1} + \delta f^s_{2n}}  \;,\\
      f^m_{2n+1^*} & = {f^s_{2n+1^*} - \delta_L f^s_{2n+1^*} + \tau \mu f^s_{2n+1}}  \;,\\
      f^m_{2n^*} & = {\mu f^s_{2n} + \delta_L f^s_{2n+1}}  \;.
    \end{aligned}
    \end{equation}
Finally, the effect of random genetic drift on genotype frequencies is given by
    \begin{equation} \label{eq:drift-single} 
      N \cdot (f'_{2n},f'_{2n+1},f'_{2n+1^*},f'_{2n^*}) \sim Mult\Big(N, (f_{2n},f_{2n+1},f_{2n+1^*},f_{2n^*})\Big),
    \end{equation}
where $Mult\Big(N, (f_{2n},f_{2n+1},f_{2n+1^*},f_{2n^*}\Big)$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies.
Overall, the change in genotype frequencies from one generation to the next is given by the transformation $f_i \to f'_i$.

\paragraph*{Multi-locus model.}
This model expands the single-locus model by allowing for the accumulation of beneficial mutations in the genome. Thus, a genotype is characterized by its karyotype, \euwt\; or \anwt, and the number of accumulated beneficial mutations, which can be zero or more.
The selection coefficient of the $i$-th mutation, $s_i$, is drawn from an exponential distribution with expected value $s$, $s_i \sim Exp(s)$.
The rest of the parameters ($N$, $\mu$, $\tau$, $\delta$, $\delta_L$, $\sigma$, $c$) are the same as in the single-locus model.
The fitness of the different genotypes is the same as in the single-locus model (\autoref{eq:fitness-single}, except that the fitness contribution of beneficial mutations is the product of all accumulated mutations, $\prod_{i \ge 0}(1+s_i)$, instead of the contribution of the single mutation allowed in the single-locus model, $(1+s)$.

@@

\paragraph*{Aneuploidy is not just another type of mutation.}
The published data indicate that, like mutation, aneuploidy can be both deleterious and beneficial~\citep{Pavelka2010, Sheltzer2011}.
Nevertheless, there are important and fundamental differences between adaptation by aneuploidy
and adaptation by beneficial mutations~\citep{Yona2015}, which make aneuploidy a unique mechanism for generating genetic
variation.
First, the aneuploidy rate (i.e. the frequency of mis-segregation events) is significantly higher than the
mutation rate~\citep{Santaguida2015review}.
Thus, everything else being equal, adaptation by aneuploidy will be faster and more frequent.
Second, fitness effects of aneuploidy are larger than those of the majority of mutations, on average, and are rarely
neutral~\citep{Pavelka2010, Yona2012, Sunshine2015}, allowing selection to quickly sort deleterious and beneficial genotypes.
Third, the number of different karyotypes is considerably smaller than the number of different genotypes, and different karyotypes are likely to have different phenotypes~\citep{Pavelka2010}.
Therefore, exploration of the phenotype space by aneuploidy requires smaller populations and a shorter time span.
Fourth, aneuploidy is a reversible state, as the rate of chromosome loss is high and the cost of aneuploidy is significant~\citep{Niwa2006}.
Indeed, aneuploidy often provides a transient solution: under short-term stress conditions, aneuploidy reverts (chromosome number returns to normal) when the stress subsides; under long-term stress conditions, aneuploidy reverts when refined solutions, generated by beneficial mutations, take over~\citep{Yona2012}.
Finally, aneuploidy results in increased genome instability, potentially increasing genetic variation by a positive feedback loop~\citep{Rancati2013, Bouchonville2009, Zhu2012}, while also increasing its own transience.

@@

\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Yitzhak Pilpel, Orna Dahan, Lilach Hadany, Judith Berman, David Gresham, Shay Covo, Martin Kupiec, and Tal Simon for discussions and comments.
This work was supported in part by 
the Israel Science Foundation (YR 552/19) and
Minerva Stiftung Center for Lab Evolution (YR).
}
